---
title: "getCensus"
author: "Eric Delmelle (with ChatGPT, and others)"
date: "2024-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install and Load Necessary Packages. 
If you haven’t already installed these packages, you'll need to do so, along with sf and ggplot2 for spatial and plotting operations.
```{r}
#install.packages(c("tidycensus", "tigris", "sf", "ggplot2", "dplyr"))
library(scales)  # Load the scales package for the rescale function
library(tidycensus)
library(tigris)
library(sf)
library(ggplot2)
library(dplyr)
library(tidyr)

```

## Set Up Census API Key.
You need to have a Census API key. If you don't have one, you can request it from the US Census Bureau. Once you have the key, you can set it in tidycensus.
```{r }
census_api_key("bd48ccbbe80006071141a76230184957dd248abe", install = TRUE,overwrite = TRUE)
```

## Fetch Census Tracts and Data. 
Choose the variables you want from the Census. You can find variable codes in the Census’s American Community Survey (ACS) documentation.

```{r}
# Fetch Census data
vars <- c("B03002_003E",  # White population
          "B19013_001E",  # Median household income
          "B15003_022E",  # Bachelor's degree or higher
          "B22001_001E",  # SNAP benefits
          "B25044_001E",  # Total households
          "B25044_003E")  # Households with no vehicle

us_data <- get_acs(geography = "tract",
                   state = state.abb[!state.abb %in% c("AK", "HI")],  # Exclude Alaska and Hawaii
                   variables = vars,
                   year = 2020,
                   survey = "acs5")

us_data <- us_data %>%
  filter(!is.na(estimate))

# Pivot data to wide format
us_data_wide <- us_data %>%
  pivot_wider(id_cols = GEOID, 
              names_from = variable, 
              values_from = estimate)

print(us_data_wide)
```

## normalize your data
```{r}
us_data_wide <- us_data_wide %>%
  mutate(across(c(B03002_003, B19013_001, B15003_022, B22001_001, B25044_001, B25044_003), 
                ~scales::rescale(.x, to = c(0, 1)))) %>%
  drop_na()

# Display the first few rows of the normalized data
head(us_data_wide)
```
## histograms of the variables
```{r}
library(ggplot2)

# Reshape the data to a long format suitable for faceting
us_data_long <- pivot_longer(us_data_wide, 
                             cols = c(B03002_003, B19013_001, B15003_022, B22001_001, B25044_001, B25044_003), 
                             names_to = "variable", values_to = "value")

# Define a vector with custom labels for the variables
variable_labels <- c(B03002_003 = "White Population",
                     B19013_001 = "Median Household Income",
                     B15003_022 = "Bachelor's Degree or Higher",
                     B22001_001 = "SNAP Benefits",
                     B25044_001 = "Total Households",
                     B25044_003 = "Households with No Vehicle")

# Create histograms with faceting
ggplot(us_data_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "#69b3a2", color = "black") +
  facet_wrap(~variable, scales = "free_x", labeller = labeller(variable = variable_labels)) +  # Use custom labels for faceting
  theme_minimal() +
  labs(title = "Distribution of Census Variables",
       x = "Value",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
# Decide how many clusters to look at
n_clusters <- 10

# Initialize total within sum of squares error: wss
wss <- numeric(n_clusters)

set.seed(123)

# Look over 1 to n possible clusters
for (i in 1:n_clusters) {
  # Fit the model: km.out
  
  km.out <- kmeans(us_data_wide[, c("B03002_003", "B19013_001", "B15003_022",
                                         "B22001_001", "B25044_003")], centers = i, nstart=20)
  # Save the within cluster sum of squares
  wss[i] <- km.out$tot.withinss
}

# Produce a scree plot
wss_df <- tibble(clusters = 1:n_clusters, wss = wss)
 
scree_plot <- ggplot(wss_df, aes(x = clusters, y = wss, group = 1)) +
    geom_point(size = 4)+
    geom_line() +
    scale_x_continuous(breaks = c(2, 4, 6, 8, 10)) +
    xlab('Number of clusters')
scree_plot

```

## kmeans
```{r}
set.seed(123)

kmeans_result <- kmeans(us_data_wide[, c("B03002_003", "B19013_001", "B15003_022",
                                         "B22001_001", "B25044_003")], centers = 10)
us_data_wide$cluster <- as.factor(kmeans_result$cluster)
```


## spatial operations
```{r}
# List of state abbreviations excluding Alaska and Hawaii
states <- state.abb[!state.abb %in% c("AK", "HI")]

# Fetch tract data for each state and transform CRS
us_tracts_list <- lapply(states, function(st) {
  tracts <- tracts(state = st, cb = TRUE)
  tracts <- st_transform(tracts, crs = 4326)  # Transform CRS to WGS84
  return(tracts)
})

# Combine all tracts into a single spatial dataframe
us_tracts <- do.call(rbind, us_tracts_list)

# Assuming us_data is already prepared and contains a column 'GEOID'
us_tracts_data <- left_join(us_tracts, us_data_wide, by = "GEOID")

```
## Plotting with GGPLOT2
```{r}
library(ggplot2)
library(sf)

# Plotting the clusters

ggplot(us_tracts_data) +
  geom_sf(aes(fill = factor(cluster)), color = NA) +  # Ensure cluster is a factor
  scale_fill_manual(values = brewer.pal(10, "Set3"), name = "Cluster") +  # Use Set3 with 10 colors
  labs(title = "K-Means Clustering of Census Data on US Conterminous Tracts",
       subtitle = "Clustered by Race, Income, and Education",
       fill = "Cluster") +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 8),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

# Assuming centroids have been calculated and stored in centroids
centroids <- kmeans_result$centers

# Define a vector with descriptive labels
# Ensure these keys EXACTLY match the column names in the centroids dataframe
descriptive_labels <- c(
  B03002_003 = "White Population",
  B19013_001 = "Median Household Income",
  B15003_022 = "Bachelor's Degree or Higher",
  B22001_001 = "SNAP Benefits",
  B25044_001 = "Total Households",
  B25044_003 = "Households with No Vehicle"
)
colnames(centroids) <- descriptive_labels[colnames(centroids)]

print(centroids)
```
## ok

```{r}

# Assuming centroids is the result of kmeans_result$centers, which might be a matrix or an array
if(is.list(centroids)) {
  # Convert list to data frame if it's indeed a list
  centroids <- as.data.frame(centroids)
}

# If centroids is a matrix (common with kmeans$centers), convert it to a dataframe
if(is.matrix(centroids)) {
  centroids <- as.data.frame(centroids)
}

# Check if the data frame conversion was successful and add the cluster column
if (is.data.frame(centroids)) {
  centroids$cluster <- as.factor(1:nrow(centroids))
} else {
  stop("Centroids is not a data frame and cannot be converted properly.")
}


# Create the bar graph
# Reshape the data to long format for ggplot
centroids_long <- pivot_longer(
  centroids,
  cols = -cluster,  # Exclude the cluster column from reshaping
  names_to = "variable",
  values_to = "value"
)

# Check the reshaped data
print(head(centroids_long))


# Create the bar graph
ggplot(centroids_long, aes(x = value, y = cluster, fill = cluster)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_wrap(~variable, scales = "free_y") +
  labs(title = "Cluster Performance on Various Variables",
       x = "Value",
       y = "Cluster") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3")  # This uses color palettes that are good for distinctions
```

```{r}
#install.packages("factoextra")
#install.packages("ggpubr")
library(factoextra)
library(ggpubr)

kmeans_result <- kmeans(us_data_wide[, c("B03002_003", "B19013_001", "B15003_022")], centers = 10)

# Perform PCA
pca_result <- prcomp(us_data_wide[, c("B03002_003", "B19013_001", "B15003_022")],
                     scale. = TRUE)  # scale. = TRUE to scale variables

# Visualize loadings
fviz_pca_var(pca_result, col.var = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE)  # Use repel = TRUE to avoid text overlapping



# Define colors from the "Set1" palette
set3_colors <- brewer.pal(10, "Set3")

# Visualize clusters using fviz_cluster with the Set1 palette
fviz_cluster(kmeans_result, data = us_data_wide[, c("B03002_003", "B19013_001", "B15003_022")],
             palette = set3_colors,  # Apply the Set1 colors
             geom = "point",
             ellipse.type = "convex",
             ggtheme = theme_bw())
```


## Interactive with leaflet
```{r}
library(leaflet)
library(RColorBrewer)  # Ensure this library is loaded for color palettes

# Creating the interactive map with a different color palette
map <- leaflet(us_tracts_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(fillColor = ~colorFactor(palette = brewer.pal(5, "Set1"), domain = us_tracts_data$cluster)(cluster),
              fillOpacity = 0.9,
              color = "#BDBDC3",
              weight = 1,
              popup = ~paste("Cluster:", cluster)) %>%
  addLegend(pal = colorFactor(palette = brewer.pal(5, "Set1"), domain = us_tracts_data$cluster),
            values = ~cluster,
            opacity = 0.8,
            title = "Cluster",
            position = "bottomright")

# Show the map
map
```


















## make sure variable name is good
```{r}
# Descriptive names for each variable
variable_names <- c(B03002_003 = "WhitePopulation",
                    B19013_001 = "MedianHouseholdIncome",
                    B15003_022 = "BachelorDegreeorHigher")

# Add a new column with descriptive names to the data frame
nc_tracts_data <- nc_tracts_data %>%
  mutate(variable_name = variable_names[variable])
```

```{r}
ggplot(nc_tracts_data) +
  geom_sf(aes(fill = normalized_estimate), color = NA) +
  facet_wrap(~variable_name, ncol = 1, labeller = label_parsed) +  # Use descriptive names
  scale_fill_viridis_c(name = "Normalized\nEstimate", option = "C") +
  theme_minimal() +
  #labs(title = "Normalized Census Data Indicators by Tract in North Carolina") +
  theme(
    strip.text = element_text(size = 8),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  )
```

## k-means
attention - only works in a wide format
```{r}

# First, pivot the data from long to wide format if not already done
library(tidyr)

nc_tracts_data <- nc_data %>%
  pivot_wider(names_from = variable, values_from = estimate)

# Rename the columns to more readable names if desired
colnames(nc_tracts_data)[colnames(nc_tracts_data) %in% c("B03002_003", "B19013_001", "B15003_022")] <- c("White_Population", "Median_Household_Income", "Bachelors_or_Higher")

# Check the first few rows to ensure correctness
head(nc_tracts_data)

```

```{r}
# Assuming no missing values, but check for them first
if (anyNA(nc_tracts_data[c("White_Population", "Median_Household_Income", "Bachelors_or_Higher")])) {
  nc_tracts_data <- na.omit(nc_tracts_data)
}

set.seed(123)  # for reproducibility
kmeans_result <- kmeans(nc_tracts_data[, c("White_Population", "Median_Household_Income", "Bachelors_or_Higher")], centers = 3)

# Add the cluster results to the data frame
nc_tracts_data$cluster <- as.factor(kmeans_result$cluster)


```




```{r}
# Assuming nc_tracts is the spatial data frame with geographic info
nc_tracts <- st_as_sf(nc_tracts)  # ensure it's an sf object

# Merge with k-means results
nc_tracts_clustered <- left_join(nc_tracts, nc_tracts_data, by = "GEOID")

# Plot the clusters
ggplot(nc_tracts_clustered) +
  geom_sf(aes(fill = cluster), color = NA) +
  scale_fill_viridis_d(name = "Cluster", discrete = TRUE) +
  labs(title = "K-Means Clustering of Census Data on NC Tracts",
       subtitle = "Clustered by Race, Income, and Education") +
  theme_void() +
  theme(
    legend.position = "right"
  )


```