---
title: "Missing and Imputing Data"
author: "EricDelmelle"
date: "2024-03-02"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Small example of code showing how to deal with missing data. This dataset contains daily air quality measurements in New York, May to September 1973.


## Load necessary libraries
```{r}
library(tidyverse)
library(ggplot2)
```

## Load the airquality dataset
```{r}
data("airquality")
head(airquality)
```

## Visualize airquality dataset
We can see the temporal variation of the airquality (Ozone here), and gaps where data is missing.
```{r}
# Adding a unique identifier for each day
airquality$DayID <- 1:nrow(airquality)

# Convert Month to a factor for better plotting
airquality$Month <- factor(airquality$Month, labels = c("May", "June", "July", "August", "September"))

# Create the spaghetti plot using ggplot
ggplot(airquality, aes(x = DayID, y = Ozone, group = Month, color = Month)) +
  geom_line() + # Add lines
  geom_point(size = 1, alpha = 0.5) + # Add points
  theme_minimal() + # Use a minimal theme
  labs(title = "Daily Ozone Levels by Month",
       x = "Day",
       y = "Ozone (ppb)",
       color = "Month") +
  theme(legend.position = "right") # Position the legend
```


## Check for missing data
```{r}
missing_data <- sapply(airquality, function(x) sum(is.na(x)))
print(missing_data)
cat("There are", sum(is.na(airquality)), "missing values in the dataset.\n")
```

## Impute missing data using the mean of the sample
Note: It's often better to consider the distribution and nature of data before imputation
```{r}
numeric_columns <- sapply(airquality, is.numeric)
airquality_imputed <- airquality
airquality_imputed[numeric_columns] <- lapply(airquality_imputed[numeric_columns], function(x) {
  if(sum(is.na(x)) > 0) {
    x[is.na(x)] <- mean(x, na.rm = TRUE)
  }
  x
})
```

## Identify outliers for the Ozone variable
```{r}
ozone <- airquality_imputed$Ozone
qnt <- quantile(ozone, probs=c(.25, .75), na.rm = TRUE)
H <- 1.5 * IQR(ozone, na.rm = TRUE)
outlier_low <- qnt[1] - H
outlier_high <- qnt[2] + H
ozone_outliers <- ifelse(ozone < outlier_low | ozone > outlier_high, TRUE, FALSE)
```

## Identify outliers
For simplicity, let's consider outliers as values more than 1.5 IQR away from the 1st and 3rd quartiles
```{r}
airquality_imputed[numeric_columns] <- lapply(airquality_imputed[numeric_columns], function(x) {
  if(sum(is.na(x)) > 0) {
    x[is.na(x)] <- mean(x, na.rm = TRUE)
  }
  x
})
```

## Prepare a dataframe for plotting
```{r}
plot_data <- data.frame(Index = 1:length(ozone), Ozone = ozone, Outlier = ozone_outliers)
```


## Spaghetti plot for Ozone variable using ggplot2, highlighting outliers
```{r}
ggplot(plot_data, aes(x = Index, y = Ozone)) +
  geom_line(color = "gray", size = 1) +  # Base line for all Ozone values
  geom_point(aes(color = Outlier), size = 2) +  # Points, colored by outlier status
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +  # Color outliers in red
  theme_minimal() +
  labs(title = "Spaghetti Plot for Ozone with Outliers",
       x = "Index",
       y = "Ozone") +
  theme(legend.position = "none")  # Remove legend for clarity
```
